---
title: "Classification Metrics"
format: html
editor: visual
---

# Classification Metrics

```{r}
#| title: add-dataset
#| echo: false
#| message: false

library(dplyr)
library(ggplot2)
library(caret)

penguin_df <- read.csv("penguin_predictions.csv")


threshold_change <- function( df, x){
  
  for(i in 1:length(df[,2])){
    if(df[i, 1] >= x){
      df[i,".pred_class"] <- "female"
    }
    
  }
  return(df)
}

penguins_CM <- function( df ){
  
  actual = factor(df[,3], levels = c("female", "male"))
  predicted = factor(df[,2], levels = c("female", "male"))
  
  confusionMatrix(predicted, actual, mode ="everything")
}
```

### Male vs Female Distribution

```{r}
#| title: null-error-rate

nr <- sum(penguin_df$sex == "female")

nullErrorRate = 1 - (nr / length(penguin_df$sex))

ggplot(data = penguin_df, aes(x = sex )) + 
  geom_bar(fill = "steelblue") 
```

Knowing the null error rate gives us a baseline understanding of whether our approach is accurate. If guessing the hypothesis every time is correct more than 95% of the time, then our model may appear to be accurate, but would never actually predict anything.

```{r}
#| title: confusion-matrix

pThresh2 <- threshold_change(penguin_df, .2)
pThresh5 <- threshold_change(penguin_df, .5)
pThresh8 <- threshold_change(penguin_df, .8)

penguins_CM(pThresh2)
penguins_CM(pThresh5)
penguins_CM(pThresh8)

```

Accuracy is noted in the above tables, Pos Pred value is the precision, recall is sensitivity, and F1 is explicit in the output.

### Threshold Use Cases

A low threshold like .2 for confusion matrices can be used when a false positive is less detrimental. If a doctor sees you have over a 20 percent chance to have a particular disease and the cost of testing is low, then a lower threshold would be preferable. A threshold of .8 would be preferred if a false positive was more detrimental. In the same scenario I would want my doctor to be more than 80 percent sure if they need to do a more invasive test with harsher consequences like a bone marrow biopsy.
