---
title: "Window Functions"
format: html
editor: visual
---

## Introduction

To start this time series analysis, I used claude to produce a random time series dataset that looks at sales of four different beverages. For the first dataset I had it look at monthly sales for simplicity, then filtered down to the year to date sales and performed a simple mean calculation for each row of the average amount sold, the average price of the beverage and the average generated revenue.

```{r}
#| title: load-data
#| message: false

df <- read.csv("monthly_sales.csv")

library(lubridate)
library(dplyr)

ytdDF <- df |>
  filter(date >= "2026-01-01")

items <- unique(df$item)
units_sold <- vector( mode = "numeric", length(items))
avg_price <- vector( mode = "numeric", length(items))
revenue <- vector( mode = "numeric", length(items))

ytdAvg <- data.frame( items = items, 
                      avg_sold = units_sold,
                      ytd_average_price = avg_price,
                      avg_rev =revenue)


for ( i in 1:length(ytdAvg$items)){
  
  tempdf <- ytdDF |>
    filter(item == ytdAvg[i,1])
  
  tempvec <- c(mean(tempdf$units_sold), mean(tempdf$avg_price), mean(tempdf$revenue))
  
  ytdAvg[i,2:4] <-tempvec
  
  rm(tempdf)
  rm(tempvec)
  
}

```

```{r}
#| title: table-1
#| echo: false

library(gt)

ytdAvg |> 
  gt() |> 
  tab_header( title = "Average sales and revenue" )

```

## monthly sales

Since I used simpler data earlier in the project, I asked claude to create another dataset that looks at the sales for this particular month and used lubridate to filter it down to last 6 days. and performed the same calculations on this dataset.

```{r}
#| title: monthly-sales 

df_2 <- read.csv("daily_sales.csv")

daily_sales_6_days <- df_2 |>
  filter( date > today() - days(6) )

rm(items, units_sold, avg_price, revenue)

items <- unique(df_2$item)
units_sold <- vector( mode = "numeric", length(items))
avg_price <- vector( mode = "numeric", length(items))
revenue <- vector( mode = "numeric", length(items))

sixDayAvg <- data.frame( items = items, 
                      avg_sold = units_sold,
                      ytd_average_price = avg_price,
                      avg_rev =revenue)

for ( i in 1:length(sixDayAvg$items)){
  
  tempdf <- daily_sales_6_days |>
    filter(item == sixDayAvg[i,1])
  
  tempvec <- c(mean(tempdf$units_sold), mean(tempdf$avg_price), mean(tempdf$revenue))
  
  sixDayAvg[i,2:4] <-tempvec
  
}

```

```{r}
#| title: table-2
#| echo: false


sixDayAvg |> 
  gt() |> 
  tab_header( title = "Average sales and revenue" )

```
